Time Series Modeling of Subdiffusive Solute Transport in an H\textsubscript{II} Phase Lyotropic Liquid Crystal Membrane

### 2019-10-09 Ben_Manuscripts/stochastic_transport/Draft.tex
Additions to intro

  Molecular modeling may make it possible to create efficient solute-specific separation
  membranes using the available chemical space of LLC monomers. To date, only a limited 
  subset of LLC monomers have been studied in membrane applications.
  ~\cite{carter_glycerol-based_2012,hatakeyama_nanoporous_2010,smith_ordered_1997,zhou_assembly_2003,resel_structural_2000}
  Even within this small subset, they have demonstrated selectivities that could not be
  explained beyond speculation and vague empirical correlations.~\cite{dischinger_application_2017}
  Atomistic molecular modeling can provide the resolution necessary to identify molecular 
  interactions that are key to separation mechanisms, allowing us to move beyond 
  Edisonian design approaches.

  In our previous work, we used molecular dynamics (MD) simulations to study the transport
  of 20 small polar molecules in an H\textsubscript{II} phase LLC membrane.~\cite{coscia_chemically_2019}
  The H\textsubscript{II} is a simpler geometry to model than the Q\textsubscript{I} phase.  
  In general, we observed subdiffusive transport behavior characterized by intermittent hops 
  separated by periods of entrapment. We identified three mechanisms responsible for the 
  solute trapping behavior: entanglement among monomer tails, hydrogen bonding with monomer
  head groups, and association with the monomer's sodium counter ions.

@@

  %MRS4: consider more citations below, when you talk about what researchers have done.
  %BJC4: added citations of prominent anomalous diffusion work
  Researchers have developed a rigorous theoretical foundation for describing the 
  motion of particles exhibiting non-Brownian, or anomalous, transport behavior.
  ~\cite{metzler_random_2000,bouchaud_anomalous_1990} The tools introduced by 
  fractional calculus are instrumental to this theory.~\cite{gorenflo_fractional_1997}
  They allow us to generalize the normally linear diffusion equation to fractional
  derivative orders, providing descriptions of a much more diverse set of behavior.~\cite{klages_anomalous_2008}

  Three well-known classes of behavior leading to anomalous subdiffusion are 
  continuous time random walks (CTRW), fractional Brownian motion
  (FBM) and random walks on fractals (RWF).\cite{meroz_toolbox_2015}
  The theory is frequently used to describe single particle trajectories.~\cite{morrin_three_2018,metzler_anomalous_2014}
  A CTRW is characterized by a distribution of hop lengths and dwell times, 
  where trajectories are characterized by independent random draws from 
  each distribution.\cite{montroll_random_1965} FBM is common in crowded,
  viscoelastic environments where each jump comes from a Gaussian distribution
  but is anti-correlated to its previous step.~\cite{mandelbrot_fractional_1968,jeon_fractional_2010,banks_anomalous_2005}
  An RWF is imposed by a system's geometry. Systems with tortuous pathways 
  and dead ends cause anti-correlated motion.\cite{meroz_toolbox_2015,neusius_subdiffusion_2008}
  These processes can occur alone or in combination.

  Our first modeling approach is based on the anomalous diffusion literature, 
  applying the mathematical formalism describing subdiffusive transport to observed
  solute trajectories. We fit anomalous diffusion models in two ways: 
  %MRS4: can you clarify the sentence below?  Probably something like: uses a single set of parameters in the diffusion equations, etc.
  %BJC4: see if below helps.
  %First, based on a single set of parameters. 
  First, we parameterized solute motion using a single set of parameters fit
  to a hypothesized anomalous diffusion model. Second, we use a two-state approach
  extracting two sets of model parameters calculated based on a solute's radial
  distance from the closest pore center.

  While elegant, pure anomalous diffusion theory may in some cases give only a 
  qualitative understanding of transport mechanisms. 
  %MRS4: might need to provide a bit more explanation why it would be ony qualitative.  
  %Relating it to molecular scale behavior and examples might help.
  %BJC4: mention ageing
  %BJC4: ?
  Complex temporal relationships in 
  between hops may have a significant effect on the average MSD. Inhomogeneity of 
  a medium or multiple types of interactions may lead to a number of distinct 
  diffusive regimes. In this case, a state based model may be useful.

@@

  Our second modeling approach applies an extended MSM framework to a relatively 
  small set of known states based on the three observed solute trapping mechanisms.  
  Standard MSMs are typically applied towards determining equilibrium populations
  of and the transitions between various state configurations.~\cite{bowman_using_2009} 
  %MRS4: this probably needs to be explained more explicitly, that the coordinates 
  %of the solute are generated according to the current state. 
  %BJC4: made more explicit.
  %We extend the framework
  %to include dynamics within states and between jumps. 
  We extend the framework to include state-dependent fluctuations and correlations in 
  solute position. 
  %MRS4: It makes sense to me what the below is, but probably isn't clear enough for others. 
  %BJC4: rephrased
%  That is we parameterize 
%  and incorporate into our model the magnitude and temporal correlation of
%  fluctuations in solute position while in each state as well as during transitions
%  to different states. 
  That is the magnitudes of a solute's fluctuations from its average trapped 
  position and the degree of correlation with its previous fluctuations is 
  defined by the parameters of its current state.

  In this work we apply the two modeling approaches in order to predict the long
  timescale behavior of the four fastest moving solutes studied in our previous work.
  Specifically, we study methanol, urea, ethylene glycol and acetic acid. In addition
  to moving quickly, allowing them to extensively explore membrane structural space,
  these solutes have a range of chemical functionality and experience each of the 
  three trapping mechanisms to different degrees. It is our goal to reproduce
  the mean squared displacements (MSDs) exhibited by each solute on MD timescales and
  then to project the long-term dynamics of each solute. We can gain further
  mechanistic insight by studying the parameters of each model. 
  %BJC3: need to decide value in long term projections

@@

  %BJC3: would take ~ 20 hours to simulation 10000 trajectories.
  %MRS4: state how much it overestimated (1% or 10000%?)
  The MSDDM over-estimates solute MSDs by 200--300 \% (see Figure~\ref{fig:msddm_performance}). 
  We simulated 1000 trajectories as described in Section~\ref{method:MSMs} then
  %BJC3: 2.2 and 3.3 x higher than MD to be exact
  calculated their MSDs. The MSD of urea and ethylene glycol are about 2 and 3 times
  higher than MD simulations respectively.

  %MRS4: It's not as clear what effect you are talking about until
  %later, i.e. the fact that the correlation is destroyed by the
  %state switching.
  The MSDDM overestimates solute MSDs because we simulated each state
  sequence independently. Although correlation exists within each
  sequence, there is no memory of the previous state or state transition 
  that can be built into the first emission of the most recent state. 
  This implies that the first value in a state sequence is a random
  draw from the underlying emission distribution. Frequent transitions
  between states lead to an abundance of short dwell times. It is 
  difficult to accurately correlate short sequences so we are 
  almost simulating uncorrelated draws from the emission distribution.
  In Figure~\ref{fig:T_sensitivity}, we demonstrate this effect by varying 
  the transition matrix and simulating a simple 2-state model with parameters
  similar to those in Table~\ref{table:msddm_params} (see Table~\ref{S-table:msddm_params}
  of the Supporting Information). We modified the probability
  transition matrix, $T$, in order to control the probability of 
  self-transitions (the diagonal elements of $T$). Larger self-transition probabilities lead to longer dwell 
  times in each state. We set the off-diagonal elements of $T$ to be equal values that make the
  rows of $T$ sum to unity when combined with the self-transition probabilities. 
  As we increase the self-transition probability, $T_{ii}$, the simulated MSD 
  decreases. Longer dwell times in each state allow us to more accurately
  simulate the negative correlation that we observe in MD.

@@

%MRS4: be more clear here what mechanistic insight it offers.
  The MSDDM offers mechanistic insight but cannot be used for reliable predictions
  of solute MSDs. It may be suitable for processes with rarer transitions between 
  states. If we can further reduce the state-space of our models, the MSDDM
  may have better predictive success, but we lose mechanistic detail.

@@

  % MRS1: will want to make a statement about whether these approaches
  % are together good enough for predictive purposes, and if not,
  % explore what else should be done in the future.
  % BJC2: It seems like it would be hard to make say these can be used
  % for prediction without experimental confirmation of the trends that we observe.

  We have tested two different mathematical frameworks for describing solute
  motion in an H\textsubscript{II} phase LLC membrane. 
  %MRS4: added some more specificity. 
  The values obtained for the parameters when fitting the models to the timeseries data  
  %The parameters of each model 
  offer important mechanistic insight on the molecular details of transport.
  Subordinated fractional Brownian and L\'evy motion have a nice theoretical 
  foundation in the anomalous diffusion literature. Our single mode model
  quantifies and allows comparison of the hopping and trapping behavior 
  between solutes. A two mode model that describes dynamics based on whether
  a solute is in or out of the pore region allows us to break down individual
  solute motion into the two distinct regimes and we showed that solute motion is
  clearly restricted while in the tail region. Our Markov state-dependent dynamical
  model uses explicitly defined trapping mechanisms and gives a nice description 
  of transitions between these observed states, the equilibrium distribution of 
  solutes among states as well as the type of stochastic behavior shown in each 
  state. However, it doesn't accurately portray correlated time series behavior
  leading to overpredicted MSDs.

@@

  Our mathematical models help us think about how to design LLC monomers for 
  solute-specific separations by forcing us to think in terms of controlling 
  solute trapping. The anomalous diffusion model is quite flexible because it
  does not require knowledge of specific trapping mechanisms. Screening a set
  of solutes and applying the anomalous diffusion model can help uncover
  trapping mechanisms by forcing the scientist to identify features common 
  to solutes with long or short hop lengths and dwell times. The MSDDM is a 
  powerful way to characterize explicit trapping mechanisms. Although its
  predictive power is inferior to the anomalous diffusion model, it clearly
  partitions a trajectory into discrete mechanisms and quantifies their
  relative dominance in the real system. For example, it is clear that 
  sodium ion association is primarily responsible for trapping urea while 
  hydrogen bonding dominates ethylene glycol. If one redesigns an LLC monomer
  to eliminate one source of trapping, 
  %MRS4: by what mechanisms are high selectivities possible?
  then high selectivities are possible.

>>>>

  Molecular modeling may make it possible to efficiently create solute-specific separation
  membranes using the available chemical space of LLC monomers. To date, only a limited 
  subset of LLC monomers have been studied in membrane applications.
  \cite{carter_glycerol-based_2012,hatakeyama_nanoporous_2010,smith_ordered_1997,zhou_assembly_2003,resel_structural_2000}
  Even within this small subset, they have demonstrated selectivities that could not be
  explained beyond speculation and vague empirical correlations.~\cite{dischinger_application_2017}
  Atomistic molecular modeling can provide the resolution necessary to identify molecular 
  interactions that are key to separation mechanisms, allowing us to move beyond 
  Edisonian design approaches.

  In our previous work, we used molecular dynamics (MD) simulations to study the transport
  of 20 small polar molecules in an H\textsubscript{II} phase LLC membrane.~\cite{coscia_chemically_2019}
  The H\textsubscript{II} phase is a simpler geometry to model than the Q\textsubscript{I} phase.  
  In general, we observed subdiffusive transport behavior characterized by intermittent hops 
  separated by periods of entrapment. We identified three mechanisms responsible for the 
  solute trapping behavior: entanglement among monomer tails, hydrogen bonding with monomer
  head groups, and association with the monomer's sodium counter ions.

@@

  %MRS4: consider more citations below, when you talk about what researchers have done.
  %BJC4: added citations of prominent anomalous diffusion work
  Researchers have developed a rigorous theoretical foundation for describing the 
  motion of particles exhibiting non-Brownian, or anomalous, transport behavior.
  \cite{metzler_random_2000,bouchaud_anomalous_1990} The tools introduced by 
  fractional calculus are instrumental to this theory.~\cite{gorenflo_fractional_1997}
  They allow us to generalize the normally linear diffusion equation to fractional
  derivative orders, providing descriptions of a much more diverse set of behavior.~\cite{klages_anomalous_2008}

  Three well-known classes of behavior leading to anomalous subdiffusion are 
  continuous time random walks (CTRW), fractional Brownian motion
  (FBM) and random walks on fractals (RWF).\cite{meroz_toolbox_2015}
  The theory is frequently used to describe single particle trajectories.~\cite{morrin_three_2018,metzler_anomalous_2014}
  %BJC4: consider rewording sentence below
  A CTRW is characterized by a distribution of hop lengths and dwell times, 
  where trajectories are characterized by independent random draws from 
  each distribution.\cite{montroll_random_1965} FBM is common in crowded,
  viscoelastic environments where each jump comes from a Gaussian distribution
  but is anti-correlated to its previous step.~\cite{mandelbrot_fractional_1968,jeon_fractional_2010,banks_anomalous_2005}
  An RWF is imposed by a system's geometry. Systems with tortuous pathways 
  and dead ends cause anti-correlated motion.\cite{meroz_toolbox_2015,neusius_subdiffusion_2008}
  These processes can occur alone or in combination.

  %BJC4: should I mention which anomalous subdiffusion model we chose? 
  Our first modeling approach is based on the anomalous diffusion literature, 
  applying the mathematical formalism describing subdiffusive transport to observed
  solute trajectories. We fit anomalous diffusion models in two ways: 
  %MRS4: can you clarify the sentence below?  Probably something like: uses a single set of parameters in the diffusion equations, etc.
  %BJC4: see if below helps.
  %First, based on a single set of parameters. 
  First, we parameterized solute motion using a single set of parameters fit
  to a hypothesized anomalous diffusion model. Second, we use a two-state approach
  extracting two sets of model parameters calculated based on a solute's radial
  distance from the closest pore center.

  %BJC4: might want to change the theme of this paragraph to something like: 
  % Anomalous diffusion can help uncover mechanisms. Once we know mechanisms, we 
  % can define them as states and use the MSDDM.
  While elegant, pure anomalous diffusion theory may in some cases give only a 
  qualitative understanding of transport mechanisms. 
  %MRS4: might need to provide a bit more explanation why it would be ony qualitative.  
  %Relating it to molecular scale behavior and examples might help.
  %BJC4: mention ageing
  %BJC4: ?
  Complex temporal relationships in 
  between hops may have a significant effect on the average MSD. Inhomogeneity of 
  a medium or multiple types of interactions may lead to a number of distinct 
  diffusive regimes. In this case, a state based model may be useful.

@@

  Our second modeling approach applies an extended MSM framework to a relatively 
  small set of known states based on the three observed solute trapping mechanisms.  
  Standard MSMs are typically applied towards determining equilibrium populations
  of and the transitions between various state configurations.~\cite{bowman_using_2009} 
  %MRS4: this probably needs to be explained more explicitly, that the coordinates 
  %of the solute are generated according to the current state. 
  %BJC4: made more explicit.
  %We extend the framework
  %to include dynamics within states and between jumps. 
  We extend the framework to include state-dependent fluctuations and correlations in 
  solute position. 
  %MRS4: It makes sense to me what the below is, but probably isn't clear enough for others. 
  %BJC4: rephrased
%  That is we parameterize 
%  and incorporate into our model the magnitude and temporal correlation of
%  fluctuations in solute position while in each state as well as during transitions
%  to different states. 
  That is the magnitudes of a solute's fluctuations from its average trapped 
  position and the degree of correlation with its previous fluctuations is 
  defined by the parameters of its current state. To distinguish our approach from
  standard MSMs, we have named it the Markov state-dependent dynamical model (MSDDM).
  %BJC4: Added name of our modeling approach in last sentence

  %BJC4: making goals more explicit
  The success of our modeling approaches can be defined in two ways. First, if we
  can closely reproduce solute MSDs measured from MD simulations with realizations
  of our models, then it is likely that the model sufficiently captures solute 
  dynamics and can be used in a predictive capacity. Even if a model fails to 
  reproduce the MD MSDs, there is value in uncovering the cause of the deviation.
  The second measure of success is by analysis of the parameters defining each 
  model. The parameters themselves should provide clues that help explain solute 
  behavior in a coherent way. A model that successfully reproduces MD MSDs, but 
  with nonsensical parameters has severely diminished value.

  % BJC4: new paragraph
  The goal of this work is not to definitively determine which modeling approach
  is better but to evaluate their performance independently because they both
  have potential value dependent on one's research goals. The anomalous subdiffusion
  approach provides a systematic way to compare the dynamical behavior of different
  solutes based solely on the time series of their center of mass positions. The process
  of choosing the correct anomalous subdiffusion provides its own mechanistic insight
  since it requires a thorough analysis of solute time series behavior. One can then
  screen different solutes, compare the fit model parameters and relate them back to 
  differences in solute size and chemical composition. The MSDDM characterizes 
  explicitly defined trapping mechanisms, providing a clear picture of solute behavior
  while in each trapped state as well as the equilibrium occupation of each trapping state.
  It is possible to use the two models in tandem, the anomalous diffusion model to
  identify mechanisms, and the MSDDM to characterize mechanisms.

  %BJC4: missing something about predictive capabilities

  We evaluate the two modeling approaches by using them to characterize the dynamical
  behavior of the four fastest moving solutes studied in our previous work.
  Specifically, we study methanol, urea, ethylene glycol and acetic acid. In addition
  to moving quickly, allowing them to extensively explore membrane structural space,
  these solutes have a range of chemical functionality and experience each of the 
  three trapping mechanisms to different extents.

%  It is our goal to reproduce
%  the mean squared displacements (MSDs) exhibited by each solute on MD timescales and
%  then to project the long-term dynamics of each solute. We can gain further
%  mechanistic insight by studying the parameters of each model. 
  %BJC3: need to decide value in long term projections

@@

  %BJC3: would take ~ 20 hours to simulation 10000 trajectories.
  %MRS4: state how much it overestimated (1% or 10000%?)
  %BJC4: added percentages. Is it redundant since I already said how many time
  The MSDDM over-estimates solute MSDs by 200--300 \% (see Figure~\ref{fig:msddm_performance}). 
  We simulated 1000 trajectories as described in Section~\ref{method:MSMs} then
  %BJC3: 2.2 and 3.3 x higher than MD to be exact
  calculated their MSDs. The MSDs of urea and ethylene glycol are about 2 and 3 times
  higher than MD simulations respectively.

  %MRS4: It's not as clear what effect you are talking about until
  %later, i.e. the fact that the correlation is destroyed by the
  %state switching.
  %BJC4: modified thesis
  The MSDDM overestimates solute MSDs because correlation is lost
  every time a state switch occurs. Although correlation exists within each
  sequence, there is no memory of the previous state or state transition 
  that can be built into the first emission of the most recent state. 
  This implies that the first value in a state sequence is a random
  draw from the underlying emission distribution. Frequent transitions
  between states lead to an abundance of short dwell times. It is 
  difficult to accurately correlate short sequences so we are 
  almost simulating uncorrelated draws from the emission distribution.
  In Figure~\ref{fig:T_sensitivity}, we demonstrate this effect by varying 
  the transition matrix and simulating a simple 2-state model with parameters
  similar to those in Table~\ref{table:msddm_params} (see Table~\ref{S-table:msddm_params}
  of the Supporting Information). We modified the probability
  transition matrix, $T$, in order to control the probability of 
  self-transitions (the diagonal elements of $T$). Larger self-transition probabilities lead to longer dwell 
  times in each state. We set the off-diagonal elements of $T$ to be equal values that make the
  rows of $T$ sum to unity when combined with the self-transition probabilities. 
  As we increase the self-transition probability, $T_{ii}$, the simulated MSD 
  decreases. Consequently longer dwell times in each state allow us to more accurately
  simulate the negative correlation that we observe in MD.

@@

  % MRS4: be more clear here what mechanistic insight it offers.
  % BJC4: added a couple sentences after thesis
  The MSDDM offers mechanistic insight but cannot be used for reliable predictions
  of solute MSDs. We can obtain a detailed picture of dynamics within each state in
  terms of the magnitude of their fluctuations and correlation with previous 
  motion. Unfortunately correlation between adjacent same-state sequences is lost
  with our model which causes the MSDs to be significantly over-predicted. The MSDDM
  may be suitable for processes with rarer transitions between states. If we can 
  further reduce the state-space of our models, it may have better predictive success,
  but we lose mechanistic detail.

@@

  We have tested two different mathematical frameworks for describing solute
  motion in an H\textsubscript{II} phase LLC membrane. The values obtained for
  the parameters when fitting the models to the time series data offer important
  mechanistic insight on the molecular details of transport.
  Subordinated fractional Brownian and L\'evy motion have a nice theoretical 
  foundation in the anomalous diffusion literature. Our single mode model
  quantifies and allows comparison of the hopping and trapping behavior 
  between solutes. A two mode model that describes dynamics based on whether
  a solute is in or out of the pore region allows us to break down individual
  solute motion into the two distinct regimes and we showed that solute motion is
  clearly restricted while in the tail region. Our Markov state-dependent dynamical
  model uses explicitly defined trapping mechanisms and gives a nice description 
  of transitions between these observed states, the equilibrium distribution of 
  solutes among states as well as the type of stochastic behavior shown in each 
  state. However, it doesn't accurately portray correlated time series behavior
  leading to overpredicted MSDs.

@@

  Our mathematical models help us think about how to design LLC monomers for 
  solute-specific separations by forcing us to think in terms of controlling 
  solute trapping. The anomalous diffusion model is quite flexible because it
  does not require knowledge of specific trapping mechanisms. Screening a set
  of solutes and applying the anomalous diffusion model can help uncover
  trapping mechanisms by forcing the scientist to identify features common 
  to solutes with long or short hop lengths and dwell times. The MSDDM is a 
  powerful way to characterize explicit trapping mechanisms. Although its
  predictive power is inferior to the anomalous diffusion model, it clearly
  partitions a trajectory into discrete mechanisms and quantifies their
  relative dominance in the real system. For example, it is clear that 
  sodium ion association is primarily responsible for trapping urea while 
  hydrogen bonding dominates ethylene glycol. If one re-designs an LLC monomer
  to eliminate one source of trapping, 
  %MRS4: by what mechanisms are high selectivities possible?
  then high selectivities are possible.
